In-silico molecular enrichment and clearance of the human intracranial space

### 2025-01-23 paper/supplementary_main.tex
Updates from Overleaf
Shuffle parts of the Discussion.

We solve the models presented in~\Cref{sec:app:more_maths} numerically
using finite element methods. The following subsections provide more
details, model-by-model.

@@

>>>>

We solve the models presented in~\Cref{sec:app:more_maths} numerically
using finite element methods. The following subsections provide more
details, model-by-model. \rami{Add a note on the convection dominated problem and the need for DG}

@@

\section{Supplementary discussion}
\label{sec:model_validation}

\subsection{Extended model validation}
\label{sec:model_validation}

In addition to the comparison of our in-silico predictions of tracer
enrichment and clearance against glymphatic MRI, we here compare
auxiliary model quantities against the literature as additional model
validation.

\paragraph{CSF flow and pressures in the SAS and ventricular system}
The dynamics of human CSF flow and pressure are better quantified, by
way of clinical imaging, in-vitro studies, and computational
modelling, in other areas of the ventricular
system~\cite{linninger2007cerebrospinal, sweetman2011cerebrospinal,
  vinje2019respiratory, hornkjol2022csf, causemann2022human,
  karki2025real, liu2025transmantle}. Linninger et
al~\cite{linninger2007cerebrospinal} model CSF flow and pressure
dynamics induced by CSF production and cardiac pulsatility under
normal and hydrocephalic conditions, and report of very good agreement
with Cine (phase-contrast) MRI measurements. Our estimates of the
maximum intracranial pressure difference, 10 Pa from the cardiac
contribution and 26 mPa from CSF production, is in perfect agreement
with their maximum transmantle pressure difference of $\sim$10 Pa, and
also in very good agreement with mean pressure differences of 11.5 Pa
measured clinically between sensors placed subdurally and in the
lateral ventricle \cite{vinje2019respiratory}. Liu et
al~\cite{liu2025transmantle} report of cardiac and respiratory
pressure differences across the aqueduct of 12.1 $\pm$ 5.7 Pa and 9.5
$\pm$ 7.2 Pa, respectively; thus our baseline estimate of the
respiratory contribution (\fixme{1.4 Pa}) may be an
underestimation. On the other hand, our cardiac- and
respiratory-driven CSF flow estimates peak at 19.8 cm/s and 4.8 cm/s
in the caudal direction, respectively, which are higher than
phase-contrast MRI measurements of cardiac and respiratory CSF flow
components \cite{takizawa2017characterization,
  yildiz2017quantifying}. Some variation in CSF flow velocities is
expected considering that the values from MRI represent averages
\cite{yildiz2017quantifying} and that the geometry of the CSF spaces
strongly affects peak velocities\cite{vinje2019respiratory,
  causemann2022human}. Hornkjøl et al \cite{hornkjol2022csf} model the
flow dynamics induced by CSF production in the choroid plexus and
report a peak CSF velocity of 8.9 mm/s in the aqueduct, which is 4.8
$\times$ higher than our values of 1.85 mm/s. Given that we use the
same production rate, this deviation again illustrates the impact of
potential differences in the (aqueduct) geometry on local velocities.

% Dispersion predictions
\paragraph{Dispersion in the SAS, ventricular system and PVS}
This pulsatile flow of CSF in the SAS, ventricular system and PVSs
leads to an increase in effective solute
diffusivity~\cite{stockman2007effect, hettiarachchi2011effect,
  asgari2016glymphatic, sharp2019dispersion, ray2021quantitative} via
a process known as Taylor dispersion \cite{taylor1953dispersion,
  watson1983diffusion}. Previous estimates of the magnitude of this
effect in the CSF spaces vary significantly: from an enhancement
factor of 0.05--1 in periarterial spaces surrounding penetrating
arteries~\cite{asgari2016glymphatic, troyetsky2021dispersion}, to
5--100 in the spinal subarachnoid space~\cite{stockman2007effect,
  hettiarachchi2011effect, sharp2019dispersion}, and up to more than
$10\,000$ in surface periarterial spaces~\cite{ray2021quantitative,
  sharp2019dispersion}. The large variability can (at least partly) be
attributed to methodological differences; e.g. different assumptions
on the medium, domain width, pressure differences and/or fluid
velocities, the diversity of CSF flow characteristics, as well as a
high likelihood of spatial variations. Hornkjøl et
al~\cite{hornkjol2022csf} consider model variations with constant
dispersion factors from 1 up to 1000, and indicate that a value of 10
gives the better agreement with the clinically observed
enrichment. Our spatially-varying estimates of the dispersion
enhancement factors $R_c, R_r$ (with $D = (1 + R_c + R_r) D^{\rm
  Gad}$) range from 0 to 200 for the cardiac contribution $R_c$ and 0
to 320 for the respiratory contribution $R_r$; and is thus compatible
within the previously reported spectrum.

% Effect of PVS size and shapes
\paragraph{Shapes, sizes and structures of the PVS}
The shapes, sizes and structures of the PVSs likely vary between
species (e.g.~mice vs.~humans), between spatial compartments
(e.g.~surface vs.~parenchymal), between vessel types (arteries
vs.~arterioles vs.~veins), and in
pathologies~\cite{ichimura1991distribution, foley2012realtime,
  schain2017cortical, mestre2018flow, bedussi2018paravascular,
  mestre2022periarteriolar, smets2024perivascular, raicevic2023sizes,
  vinje2021brain, eide2024functional}. In terms of shape, the PVSs are
commonly represented as annular (elliptic) cylinders, though it is
well recognized that this represents an
idealization~\cite{mestre2018flow, tithof2019hydraulic,
  vinje2021brain, raicevic2023sizes, boster2024hydraulic,
  smets2024perivascular}. In terms of sizes, Raicevic et
al~\cite{raicevic2023sizes} note that the variation in PVS area is
larger between PVS segments than along a single PVS segment and that
the PVS area increases with lumen area. In mice, reports of the ratio
between PVS and lumen area range from
$\approx$0.35--0.43~\cite{smets2024perivascular} up to
$\approx$1.12--1.4~\cite{raicevic2023sizes, mestre2018flow}. In
humans, the PVS may be as wide as the associated surface artery and up
to 4$\times$ wider in iNPH subjects~\cite{eide2024functional}, which
would correspond to substantially larger PVS area ratios (3 or
higher). To reflect the human scale, we here represent each PVS
segment as an annular cylinder with inner radius $R_1$ and outer
radius $R_2$ of width and area proportional to that of the
corresponding blood vessel ($R_2 = 2 R_1$ at baseline, $R_2 = 3 R_1$
for enlarged PVS). The hydraulic resistance of annular cross-sections
is $1-6 \times$~\cite{tithof2019hydraulic} larger than more elongated
cross-sections and thus our estimates of the pressure-induced PVS
velocities are conservative.
